PyPy's Approach to Virtual Machine Construction

### 2006-05-31 talk/dls2006/paper.tex
More URL fighting.
title
sort out quote marks (I love query-replace-regexp)

\title{Still Missing a Cool Title}

% title ideas: Implementing Virtual Machines in Dynamic Languages?

@@

For this and other reasons, standard platforms emerge; nowadays, a
language implementer could cover most general platforms in use by
writing three versions of his virtual machine: for C/Posix, for Java,
and for CLI/.NET.  This is, at least, the current situation of the
Python programming language, where independent volunteers have
developed and are now maintaining Java and .NET versions of Python,
which follow the evolution of the "official" C version (CPython).

However, we believe that platform standardization does not have to be
a necessary component of this equation.  We are basically using the
standard "meta-programming" argument: if one could write the VM in a
very high level language, then the VM itself could be automatically
\textit{translated} to any lower-level platform.  Moreover by writing
the VM in such a language we would gain in flexibility in
architectural choices and expressiveness.

@@

There are two major components in PyPy:
%
\begin{enumerate}
\item the \textit{Standard Interpreter}: an implementation of the Python programming
language, mostly complete and compliant with the current version of the
language, Python 2.4.
\item the \textit{Translation Process}: a translation tool-suite whose goal is to
compile subsets of Python to various environment.
\end{enumerate}
%
In particular, we have defined a subset of the Python language called
"restricted Python" or RPython.  This sublanguage is not restricted
syntactically, but only in the way it manipulates objects of different
types.  The restrictions are a compromise between the expressivity and
the need to statically infer enough types to generate efficient code.
The foremost purpose of the translation tool-suite is to compile such
RPython programs to a variety of different platforms.

@@

This means that in addition to transforming the existing graphs, each
transformation step also needs to insert new functions into the forest.
A key feature of our approach is that we can write such "system-level"
code -- relevant only to a particular transformation -- in plain Python
as well.  The idea is to feed these new Python functions into the
front-end, using this time the transformation's target (lower-level)
type system during the type inference.  In other words, we can write
plain Python code that manipulates objects that conform to the
lower-level type system, and have these functions automatically
transformed into appropriately typed graphs.

@@

In the example of the \texttt{malloc} operation, replaced by a call to
GC code, this GC code can invoke a complete collection of dead
objects, and can thus be arbitrarily complicated.  Still, our GC code
is entirely written in plain Python, and it manipulates "objects" that
are still at a lower level: pointer and address objects.  Even with
the restriction of having to use pointer-like and address-like
objects, Python remains more expressive than, say, C to write a GC
(Jikes RVM's GC work \cite{JikesGC} was the inspiration to try to
express GCs in Python, see section \ref{relatedwork}).

@@

Now, clearly, the purpose of types like a "C-like struct" or a "C-like
array" is to be translated to a real \texttt{struct} or array declaration by
the C back-end.  What, then, is the purpose of emulating such things in
Python?  The answer is three-fold.  Firstly, if we have objects that
live within the Python interpreter, but faithfully emulate the behavior
of their C equivalent while performing additional safety checks, they
are an invaluable help for testing and debugging.  For example, we can
check the correctness of our hash table implementation, written in
Python in term of struct- and array-like objects, just by running it.
The same holds for the GC.

@@

Note that the left column, i.e.\ the bytecode interpreter and the
Standard Object Space, form the full Python interpreter of PyPy.  It is
an RPython program, and the whole purpose of the translation process is
to accept this as \textit{input}, and translate it to an efficient form.
The description of this particular input program is beyond the scope of
the present paper; see [S].

@@

The Flow Object Space records all operations that the bytecode
interpreter "would like" to do between the placeholder objects.  It
records them into basic block objects that will eventually be part of
the control flow graph of the whole function.  The recorded operations
take Variables and Constants as argument, and produce new Variables as
results.  The Constants serve two purposes: they are a way to
introduce constant values into the flow graphs -- these values may be
arbitrarily complex objects, not just primitives -- and they allow
basic constant propagation.\footnote{This is useful at this level for
some constructs of the bytecode interpreter, which can temporarily
wrap internal values and push them onto the regular value stack among
the other application-level objects.  We need to be able to unwrap
them again later.  Moreover, we rely on this feature to perform
compile-time computations, particularly in the generic system code
helpers, which ask for and compute with the concrete types of the
variables they receive.}

@@

Here is a simplified, static model of how the annotator works.  It can
be considered as taking as input a finite family of functions calling
each other, and working on the control flow graphs of each of these
functions as built by the Flow Object Space (section \ref{flowobjspace}).
Additionally, for a particular "entry point" function, the annotator
is provided with user-specified types for the function's arguments.

@@

We can consider that all variables are initially assigned the "bottom"
type corresponding to the empty set of possible run-time values.  Types
can only ever be generalised, and the model is simple enough to show
that there is no infinite chain of generalisation, so that this process
necessarily terminates.

@@

The set $A$ of RPython types is defined as the following formal terms:
%
\begin{itemize}
\item $Bot$, $Top$ -- the minimum and maximum elements (corresponding
      to "impossible value" and "most general value");

@@

The type system moreover comes with a family of rules, which for every
operation and every sensible combination of input types describes the
type of its result variable.  Let $V$ be the set of Variables that
appear in the user program's flow graphs.  Let $b$ be a map from $V$
to $A$; it is a "binding" that gives to each variable a type.  The
purpose of the annotator is to compute such a binding stepwise.

@@

{\bf pypy-c-stackless.}
    The same as pypy-c, plus the "stackless transformation" step which
    modifies the flow graph of all functions in a way that allows them
    to save and restore their local state, as a way to enable coroutines.

{\bf pypy-c-gcframework.}
    In this variant, the "gc transformation" step inserts explicit
    memory management and a simple mark-and-sweep GC implementation.
    The resulting program is not linked with Boehm.  Note that it is not
    possible to find all roots from the C stack in portable C; instead,
    in this variant each function explicitly pushes and pops all roots
    to an alternate stack around each subcall.

{\bf pypy-c-stackless-gcframework.}
    This variant combines the "gc transformation" step with the
    "stackless transformation" step.  The overhead introduced by the
    stackless feature is theoretically balanced with the removal of the
    overhead of pushing and popping roots explicitly on an alternate
    stack: indeed, in this variant it is possible to ask the functions
    in the current C call chain to save their local state and return.
    This has the side-effect of moving all roots to the heap, where the
    GC can find them.  (We hypothesize that the large slowdown is caused
    by the extreme size of the executable in this case -- 21MB, compared to
    6MB for the basic pypy-c.  Making it smaller is work in progress.)

@@

Other transformations like the "gc" and the "stackless" ones actually
take more time, particuarly when used in combination with each other (we
speculate it is because of the increase in size caused by the previous
transformations).  A translation of pypy-c-stackless, without counting
GCC time, takes 60 minutes; the same for pypy-c-stackless-gcframework
takes 129 minutes.

@@

So far, the PyPy tool-chain can only translate the Python interpreter of
PyPy into a program which is again an interpreter -- the same interpreter
translated to C, essentially, although we have already shown that some
aspects can be "weaved" in at translation time, like support for
coroutines.

@@

Squeak \cite{Squeak} is a Smalltalk implementation in Smalltalk. It uses SLang, a
very restricted subset of Smalltalk with few types and strict
conventions, which can be mostly directly translated to C.
The VM, the object memory and the garbage collector support are
explicitly written together in this style. Again simplicity and
portability were the major goals, as opposed to sophisticated manipulation
and analysis or "weaving" in of features as transformation aspects.

>>>>

\title{PyPy's Approach to Virtual Machine Construction}

@@

For this and other reasons, standard platforms emerge; nowadays, a
language implementer could cover most general platforms in use by
writing three versions of his virtual machine: for C/Posix, for Java,
and for CLI/.NET.  This is, at least, the current situation of the
Python programming language, where independent volunteers have
developed and are now maintaining Java and .NET versions of Python,
which follow the evolution of the ``official'' C version (CPython).

However, we believe that platform standardization does not have to be
a necessary component of this equation.  We are basically using the
standard ``meta-programming'' argument: if one could write the VM in a
very high level language, then the VM itself could be automatically
\textit{translated} to any lower-level platform.  Moreover by writing
the VM in such a language we would gain in flexibility in
architectural choices and expressiveness.

@@

There are two major components in PyPy:
%
\begin{enumerate}
\item the \textit{Standard Interpreter}: an implementation of the Python programming
language, mostly complete and compliant with the current version of the
language, Python 2.4.
\item the \textit{Translation Process}: a translation tool-suite whose goal is to
compile subsets of Python to various environment.
\end{enumerate}
%
In particular, we have defined a subset of the Python language called
``restricted Python'' or RPython.  This sublanguage is not restricted
syntactically, but only in the way it manipulates objects of different
types.  The restrictions are a compromise between the expressivity and
the need to statically infer enough types to generate efficient code.
The foremost purpose of the translation tool-suite is to compile such
RPython programs to a variety of different platforms.

@@

This means that in addition to transforming the existing graphs, each
transformation step also needs to insert new functions into the forest.
A key feature of our approach is that we can write such ``system-level''
code -- relevant only to a particular transformation -- in plain Python
as well.  The idea is to feed these new Python functions into the
front-end, using this time the transformation's target (lower-level)
type system during the type inference.  In other words, we can write
plain Python code that manipulates objects that conform to the
lower-level type system, and have these functions automatically
transformed into appropriately typed graphs.

@@

In the example of the \texttt{malloc} operation, replaced by a call to
GC code, this GC code can invoke a complete collection of dead
objects, and can thus be arbitrarily complicated.  Still, our GC code
is entirely written in plain Python, and it manipulates ``objects'' that
are still at a lower level: pointer and address objects.  Even with
the restriction of having to use pointer-like and address-like
objects, Python remains more expressive than, say, C to write a GC
(Jikes RVM's GC work \cite{JikesGC} was the inspiration to try to
express GCs in Python, see section \ref{relatedwork}).

@@

Now, clearly, the purpose of types like a ``C-like struct'' or a ``C-like
array'' is to be translated to a real \texttt{struct} or array declaration by
the C back-end.  What, then, is the purpose of emulating such things in
Python?  The answer is three-fold.  Firstly, if we have objects that
live within the Python interpreter, but faithfully emulate the behavior
of their C equivalent while performing additional safety checks, they
are an invaluable help for testing and debugging.  For example, we can
check the correctness of our hash table implementation, written in
Python in term of struct- and array-like objects, just by running it.
The same holds for the GC.

@@

Note that the left column, i.e.\ the bytecode interpreter and the
Standard Object Space, form the full Python interpreter of PyPy.  It is
an RPython program, and the whole purpose of the translation process is
to accept this as \textit{input}, and translate it to an efficient form.
The description of this particular input program is beyond the scope of
the present paper; see \cite{S}.

@@

The Flow Object Space records all operations that the bytecode
interpreter ``would like'' to do between the placeholder objects.  It
records them into basic block objects that will eventually be part of
the control flow graph of the whole function.  The recorded operations
take Variables and Constants as argument, and produce new Variables as
results.  The Constants serve two purposes: they are a way to
introduce constant values into the flow graphs -- these values may be
arbitrarily complex objects, not just primitives -- and they allow
basic constant propagation.\footnote{This is useful at this level for
some constructs of the bytecode interpreter, which can temporarily
wrap internal values and push them onto the regular value stack among
the other application-level objects.  We need to be able to unwrap
them again later.  Moreover, we rely on this feature to perform
compile-time computations, particularly in the generic system code
helpers, which ask for and compute with the concrete types of the
variables they receive.}

@@

Here is a simplified, static model of how the annotator works.  It can
be considered as taking as input a finite family of functions calling
each other, and working on the control flow graphs of each of these
functions as built by the Flow Object Space (section \ref{flowobjspace}).
Additionally, for a particular ``entry point'' function, the annotator
is provided with user-specified types for the function's arguments.

@@

We can consider that all variables are initially assigned the ``bottom''
type corresponding to the empty set of possible run-time values.  Types
can only ever be generalised, and the model is simple enough to show
that there is no infinite chain of generalisation, so that this process
necessarily terminates.

@@

The set $A$ of RPython types is defined as the following formal terms:
%
\begin{itemize}
\item $Bot$, $Top$ -- the minimum and maximum elements (corresponding
      to ``impossible value'' and ``most general value'');

@@

The type system moreover comes with a family of rules, which for every
operation and every sensible combination of input types describes the
type of its result variable.  Let $V$ be the set of Variables that
appear in the user program's flow graphs.  Let $b$ be a map from $V$
to $A$; it is a ``binding'' that gives to each variable a type.  The
purpose of the annotator is to compute such a binding stepwise.

@@

{\bf pypy-c-stackless.}
    The same as pypy-c, plus the ``stackless transformation'' step which
    modifies the flow graph of all functions in a way that allows them
    to save and restore their local state, as a way to enable coroutines.

{\bf pypy-c-gcframework.}
    In this variant, the ``gc transformation'' step inserts explicit
    memory management and a simple mark-and-sweep GC implementation.
    The resulting program is not linked with Boehm.  Note that it is not
    possible to find all roots from the C stack in portable C; instead,
    in this variant each function explicitly pushes and pops all roots
    to an alternate stack around each subcall.

{\bf pypy-c-stackless-gcframework.}
    This variant combines the ``gc transformation'' step with the
    ``stackless transformation'' step.  The overhead introduced by the
    stackless feature is theoretically balanced with the removal of the
    overhead of pushing and popping roots explicitly on an alternate
    stack: indeed, in this variant it is possible to ask the functions
    in the current C call chain to save their local state and return.
    This has the side-effect of moving all roots to the heap, where the
    GC can find them.  (We hypothesize that the large slowdown is caused
    by the extreme size of the executable in this case -- 21MB, compared to
    6MB for the basic pypy-c.  Making it smaller is work in progress.)

@@

Other transformations like the ``gc'' and the ``stackless'' ones actually
take more time, particuarly when used in combination with each other (we
speculate it is because of the increase in size caused by the previous
transformations).  A translation of pypy-c-stackless, without counting
GCC time, takes 60 minutes; the same for pypy-c-stackless-gcframework
takes 129 minutes.

@@

So far, the PyPy tool-chain can only translate the Python interpreter of
PyPy into a program which is again an interpreter -- the same interpreter
translated to C, essentially, although we have already shown that some
aspects can be ``weaved'' in at translation time, like support for
coroutines.

@@

Squeak \cite{Squeak} is a Smalltalk implementation in Smalltalk. It uses SLang, a
very restricted subset of Smalltalk with few types and strict
conventions, which can be mostly directly translated to C.
The VM, the object memory and the garbage collector support are
explicitly written together in this style. Again simplicity and
portability were the major goals, as opposed to sophisticated manipulation
and analysis or ``weaving'' in of features as transformation aspects.
